Gradual Soundness: Lessons from Static Python

### 2021-12-06 pj-7.1/paper.tex

\begin{itemize}
  \item
    Evidence that gradual soundness delivers Python-level performance at a low migration cost
    and faster performance as programmers gradually incorporate sound types.
    \SP{} can be applied to a codebase with no refactoring and marginal performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description of the \SP{} core language and an analysis of its soundness
    and optimizations.
    Soundness is not uniform across all types.
    Some are fully sound while others are constructor sound.
  \item
    Discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

The designers of Static Python hope to eventually build a system that is sound for arbitrary Python code,
but are not there yet.
The known side channels include:
eval, the locals and globals dicts (the latter can be accessed through frame objects at the moment),
and mutable closure cells.
Our model does not cover these either.
We essentially ban all these features, rejecting any program in which they appear.
Code mining in JS (and perhaps the newer R work?) suggests that at least eval
uses can be rewritten without too much pain. The few exceptions would point
to language weaknesses / call for new semantically-descriptive features.

FILL do we stop at strict modules?\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
A strict module is 100\% repeatable and isolated from the side effects of other modules.

@@

@@

>>>>

\begin{itemize}
  \item
    Evidence that gradual soundness delivers Python-level performance at a low migration cost
    and faster performance as programmers gradually incorporate sound types.
    \SP{} can be applied to a codebase with no refactoring and marginal performance overhead.
    After refactoring to sound types, \SP{} performs well on a microbenchmark
    suite and in production.
  \item
    A formal description of the \SP{} core language and an analysis of its soundness
    and optimizations.
    Soundness is gradual in the sense that some types are fully sound and others validate
    only top-level shapes.
  \item
    Discussions of the expressiveness limitations in \SP{} relative to
    idealized gradual type systems, and the programming discipline that Instagram
    follows in their gradually-sound codebase.
\end{itemize}

@@

Our model of \SP{} includes the key aspects of its types,
statements, and values.
There are two main goals:
illustrate the boundaries where typed and untyped code can mix,
and to show how to soundly protect these boundaries.
Compared to other gradual languages, these boundaries are quite restrictive---which
partly explains \SP{} performance~(\cref{s:eval}).
Indeed the model is nearly a subset of model for Nom~\cite{mt-oopsla-2021},
hence we focus on explaining the gaps and omit formal proofs.

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle soundly (FILL examples ... while loop?);
omitting these lets us focus on the difficult corners.
Other Python features are left out because \SP{} does not ascribe types
to them.
These include \code{eval}, first-class classes, and module-level cells;
see \cref{s:impl} for further discussion.

@@

%% SP side channels:
%% eval,
%% the locals and globals dicts (the latter can be accessed through frame objects at the moment),
%% and mutable closure cells.

@@

\subsection{Dynamic Python Features}

\SP{} does not ascribe types to the following Python features.
These are not covered in our model because the implementation simply
gives them the dynamic type and lets the runtime treat them as untyped
Python code.
For each one, the dynamic type is a reasonable choice; static types are
likely to cause more trouble than helpful structure.

\paragraph{First-Class Classes}

\SP{} does not attempt to type first-class classes:
partly because they see little use at Instagram (FILL how much?)
and partly because it is unclear how to incorporate them into
the nominal type system.
The straightforward but restrictive approach is to force code
that uses a first-class class to expect instances of a particular
static class.
FILL is the bracha cook mixin pattern a way to work within nominal~\cite{bc-oopsla-1990}?
\cite{fkf-popl-1998} propose a different method that accommodates mixins,
but the method calls a second layer of interface types.
\cite{mt-oopsla-2021} similarly allow structural objects with interfaces.
Adding these layers to \SP{} is an open question; in particular,
prior work does not handle multiple inheritance.

%First-class classes are untyped
%(good discussion here:
%- mixin based tends to structrural types
%- nominality overly restrictive, get forced into <: hierarchy
%- but don't have NO IDEA for wat to do
%  + classes and mixins paper shows one idea,
%    ... need complicated mixin study??? (many sub objects)
%    built new type system to be mixin-aware, layered atop java
%    unsure if such extension works for SP
%    furthermor, class-and-mixins depends on Java single inheritance
%  + bracha cook mixin pattern
%  + tate muehlboeck structural object + interfaces

\paragraph{Dynamic Execution}
%% TODO is eval a side channel, or safe?
%% TODO check that eval/exec args really are unoptimized / sound etc.
Results computed by calls to \code{eval} and \code{exec}
have the dynamic type.
Their inputs run without being rewritten
by the optimizer.

Studies of JavaScript and R have shown that many uses of dynamic execution
can be removed through simple adjustments~\cite{rhbv-ecoop-2011,gdkkv-oopsla-2021,mrmv-esop-2012}.
Assuming these findings carry over to Python, we recommend either similar
adjustments or the introduction of semantically-descriptive replacements
over attempting to type eval.

\paragraph{Module-Level Variables}

Any Python module can read and write to the module-level variables of another module.
When compiling a module, \SP{} therefore assumes that its module-level variables
may be modified by untyped code and assigns the dynamic type.

Cinder offers \emph{strict modules} as an alternative to the Python
semantics.\footnote{https://instagram-engineering.com/python-at-scale-strict-modules-c0bb9245c834}
If a programmer chooses to declare a module as strict, then its module-level varibles
are immutable and thus typeable.

Another potential solution is for the Cinder runtime to check that writes to module-level variables
preserve their types.
There are two downsides to this idea:
it will add some performance overhead,
and it will force developers to rewrite untyped code in order to fix any errors that arise.
%% Because of the latter concern, type dynamic is a very reasonable default.
